---
title: "Disaggregation of ASA24 Foods"
author: "Stephanie Wilson"
date: "February 2026"
output:
  md_document:
    variant: gfm
    toc: true
    number_sections: false
    df_print: kable
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    theme: sandstone
    highlight: tango
    df_print: paged
---

## Disaggregation of ASA24 Foods
This script takes in ASA24 ITEMS files, disaggregates WWEIA food codes to ingredients, and calculates the new ingredient weight. This script also calculates total caloric intake & other nutrients for each participant recall so polyphenol intakes can be standardized to caloric intake later on. 

#### INPUTS

-   **Your Dietary Data** - This script does not provide filtering for portion or nutrient outliers. These may be performed in advance. The ASA24 website has cleaning recommendations here: ["Reviewing and Cleaning ASA24 Data"](https://epi.grants.cancer.gov/asa24/resources/cleaning.html).

-   **FDA_FDD_All_Records_v_3.1.xlsx** - FDD FoodCodes to Ingredients and Ingredient Percentages

#### OUTPUTS

-   **Diet_Disaggregated.csv.bz2**: Dietary data that has been disaggregated using FDD.
-   **Diet_total_nutrients.csv**: Total daily kcal intakes for unique records (subject, RecallNo for Recall; subject, RecordNo, RecordDayNo for Record)

## SCRIPT

```{r echo = FALSE, purl=FALSE}
# Clears all objects from memory
rm(list = ls())
```

```{r include=FALSE}
knitr::opts_knit$set(root.dir = "..") # project root relative to scripts/
```

```{r warning = FALSE}
# Load packages
suppressMessages(library(dplyr))
suppressMessages(library(vroom))
suppressMessages(library(tidyr))
suppressMessages(library(stringr))
suppressMessages(library(readxl))
```

Load Example Dietary Data and FDA-FDD V3.6
```{r}
# Load provided file paths
source("provided_files.R")

# Load User Dietary Data
input_data = vroom::vroom(diet_input_file, show_col_types = FALSE) %>%
  dplyr::rename(subject = UserName)

# FDD Disaggregation options
# Rename for ease of use.
FDD_V3 = readxl::read_xlsx(FDD_file) %>%
  dplyr::rename(latest_survey = "Latest Survey",
         wweia_food_code = "WWEIA Food Code",
         wweia_food_description = "WWEIA Food Description",
         fdd_ingredient = "Basic Ingredient Description",
         ingredient_percent = "Ingredient Percent") %>%
  dplyr::select(wweia_food_code, wweia_food_description,  
         fdd_ingredient, ingredient_percent) %>%
  dplyr::mutate(wweia_food_code = as.integer(wweia_food_code))
```

### Dietary Data Filtering

Filter IN Individuals With More than Recall/Record and filter OUT incomplete recalls/Records

-   RecallStatus: 2=Complete; 5=Breakoff/Quit
-   RecordDayStatus: 2=Complete; 5=Breakoff/Quit

```{r}
if ("RecallNo" %in% names(input_data)) {
  
  # Recall dataset
  input_data_clean = input_data %>%
    group_by(subject) %>%
    filter(n_distinct(RecallNo) > 1) %>%
    ungroup() %>%
    filter(RecallStatus != 5)
  
} else if ("RecordNo" %in% names(input_data)) {
  
  # Record dataset
  input_data_clean = input_data %>%
    group_by(subject) %>%
    filter(n_distinct(RecordNo) > 1 | n_distinct(RecordDayNo) > 1) %>%
    ungroup() %>%
    filter(RecordDayStatus != 5)
  
} else {
  stop("Data must contain RecallNo or RecordNo.")
}
```

### Specify grouping variables
Column grouping depends on whether ASA24 output is from a record or recall.
```{r}
if ("RecallNo" %in% names(input_data_clean)) {
  group_vars = c("subject", "RecallNo")
  
} else if ("RecordNo" %in% names(input_data_clean)) {
  group_vars = c("subject", "RecordNo", "RecordDayNo")
  
} else {
  stop("Data must contain RecallNo or RecordNo.")
}
```

### Sum by Entry (Recall/Record) for total kcal and other nutrient totals

```{r}
input_total_nutrients = input_data_clean %>%
  dplyr::group_by(across(all_of(group_vars))) %>%
  dplyr::summarize(across(KCAL:B12_ADD, ~ sum(.x, na.rm = TRUE), 
                          .names = "Total_{.col}")) %>%
  dplyr::ungroup()
```

### Minimize the number of columns to the essential data

```{r}
input_data_clean_minimal = input_data_clean %>%
  dplyr::rename(wweia_food_code = FoodCode,
         food_description = Food_Description) %>%
  dplyr::select(c(subject,
                  any_of(c("RecallNo", "RecordNo", "RecordDayNo")),
                  wweia_food_code, food_description, FoodAmt))
```

### Apply Ingredient Percentage Adjustment for Coffee and Tea Brewing    
```{r}
FDD_V3_adjusted = FDD_V3 %>%
  dplyr::group_by(wweia_food_code) %>%
  dplyr::mutate(
    
    # Create Flag 
    has_tea    = any(str_detect(fdd_ingredient, regex("Tea", ignore_case = TRUE))),
    has_coffee = any(str_detect(fdd_ingredient, regex("Coffee", ignore_case = TRUE))),
    has_water  = any(str_detect(fdd_ingredient, regex("Water", ignore_case = TRUE))),

    # Add combined coffee|tea + water percentages
    brewing_adjustment_total = case_when(
      # Tea + Water
      has_tea & has_water ~ sum(
        ingredient_percent[str_detect(fdd_ingredient, regex("Tea|Water", ignore_case = TRUE))],
        na.rm = TRUE),
      # Coffee + Water
      has_coffee & has_water ~ sum(
        ingredient_percent[str_detect(fdd_ingredient, regex("Coffee|Water", ignore_case = TRUE))],
        na.rm = TRUE),
      TRUE ~ NA_real_),
    
    # Ensure new brewing adjustment percentage is applied to only coffee|tea
    brewing_adjustment_percentage = if_else(
      str_detect(fdd_ingredient, regex("Coffee|Tea", ignore_case = TRUE)),
      brewing_adjustment_total,
      NA_real_)) %>%
  dplyr::select(-c(has_tea, has_coffee, has_water, brewing_adjustment_total)) %>%
  dplyr::ungroup()
```


### Disaggregate Food Codes and compute final Ingredient Weights

```{r}
merge = dplyr::left_join(input_data_clean_minimal, FDD_V3_adjusted, by = "wweia_food_code", 
                  relationship = "many-to-many") %>%
  # Compute final ingredient weight
  # If brewing adjustment exists, it will use this adjustment first.
  dplyr::mutate(FoodAmt_Ing_g = FoodAmt * (
      coalesce(brewing_adjustment_percentage, ingredient_percent) / 100))
```


### Write output files
Ensure outputs directory is created
```{r}
if (!dir.exists("outputs")) dir.create("outputs", recursive = TRUE)
```

Write Files
```{r}
vroom::vroom_write(merge, 'outputs/Diet_Disaggregated.csv.bz2', delim = ",")
vroom::vroom_write(input_total_nutrients, 'outputs/Diet_total_nutrients.csv', delim = ",")
```
